---
title: <br> 
  Final Report
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="75")
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	prompt = FALSE,
	results = "hide"
)
opts_knit$set(width=75)
```

# 1. Summary

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are tracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

Currently, the Cyclistic offers three different kinds of memberships :

*single-ride passes (casual riders)
* full-day passes (casual riders)
* annual memberships (Cyclistic riders)

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, the team believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers, they believe there is a very good chance to convert casual riders into members.

This report will assess and design marketing strategies aimed at converting casual riders into annual members.

# 2. Prepare and Process

For the purpose of this analysis, only data between September 2020 and August 2021 provided by the Cyclistic will be assessed. Then the data has been processed through numerous steps which can be seen under the “Data Cleaning” tab. 

Due to the large file sizes, R has been used as the primary tool to clean and process the datasets. There is minimal human error and data bias since the data is taken from the bikes themselves being geo tracked. However, due to data privacy rules, there is no data relating to the type of user.

However, note that because of data-privacy issues no data relating to riders’ personally identifiable information is used. This means that no connection between pass purchases to credit card numbers to determine if casual riders live in the Cyclistic service area or if they have purchased multiple single passes is done.

The data has been cleaned by way of merging all 12 datasets into one, deleting incomplete data elements, removing test station results, removing negative ride lengths and summarising the dataset by date and time variables. The full data cleaning process has been documented in the “Data Cleaning ” tab.

The data cleaning process highlighted that :

* the “classic” bike type was added to Cyclistic’s portfolio from December 2020 onwards which can be seen under the “Data Cleaning tab” followed by “Understanding the datasets” followed by “Checking Rideable types”.
* that there are a few stations that have been added or removed from Cyclistic’s portfolio which can be seen under the “Data Cleaning tab” followed by “Understanding the datasets” followed by “Checking the station name”.

Finally, the cleaned dataset has been saved under the file name “all_trips_cleaned”.

```{r echo=FALSE}
library(tidyverse)
library(data.table)
library(viridis)
library(ggpubr)
```

```{r echo=FALSE}
all_trips_cleaned <- fread("C:/Users/nayya/OneDrive/Desktop/Portfolio/Cyclistic/Raw Data//all_trips_cleaned.csv")
```

# 3. Analysis

### 3.1 Most popular station

```{r echo=FALSE}
library(leaflet)
library(htmlwidgets)
library(htmltools)
```

```{r echo=FALSE}
map_data <- all_trips_cleaned %>% 
  select(start_station_name, start_lat, start_lng) %>% 
  group_by(start_station_name) %>% 
  mutate(numtrips = n()) %>% 
  distinct(start_station_name, .keep_all = TRUE)

mybins <- seq(0, 70000, by = 10000)

mypalette <- colorBin(palette = "viridis", domain = map_data$numtrips,
                      na.color = "transparent", bins = mybins)

mytext <- paste("Start station name :", map_data$start_station_name, "<br/>",
                "Number of trips :", map_data$numtrips, sep = "") %>% 
  lapply(htmltools::HTML)
```

```{r echo=FALSE}
plot_1 <- leaflet(map_data) %>% 
  addTiles() %>% 
  setView(lng = -87.6298, lat = 41.8781, zoom = 11.5) %>% 
  addProviderTiles("Esri.WorldGrayCanvas") %>% 
  addCircleMarkers(~start_lng, ~start_lat, fillColor = ~mypalette(numtrips),
                   fillOpacity = 0.7, color = "white", radius = 8, stroke = FALSE,
                   labelOptions = labelOptions(style = list("font-weight"="normal",
                                                            padding = "3px 8px"),
                                               textsize = "13px",
                                               direction = "auto")) %>% 
  addLegend(pal = mypalette, values = ~numtrips, opacity = 0.9,
            title = "Number of trips", position = "bottomright")
```

```{r interactive map, results = "show"}
plot_1
```

From the interactive map the following conclusions can be made :

* The shoreline bike stations centred around the Navy Pier are the most popular, with Streeter Dr & Grand Ave being Cyclistic’s most popular station with 64,998 trips, confirming Cyclistic’s internal analysis that the majority of its users ride for leisure.

* Interacting with the map highlights that the southern stations are Cyclistic’s least popular stations despite the stations covering a large surface area. This indicates that only a small proportion of Cyclistic users use the bikes to commute to and from more residential areas, and instead the vast majority of users use the bikes around the central, tourist locations of Chicago.

### 3.2 Trip duration

```{r echo=FALSE}
plot_2 <- ggplot(data=all_trips_cleaned, aes(ride_length, fill=factor(member_casual))) +
  geom_histogram(binwidth = 300) + 
  xlim(0, 3600) +
  scale_y_continuous(labels = scales::label_number_si()) +
  facet_wrap(~member_casual) +
  scale_fill_viridis_d()
```

```{r echo=FALSE}
plot_2
```

The above-mentioned histogram demonstrates that the durationsof trips are fairly similar for both casual and member trips, peaking somewhere around the 10-15 minute mark. These seem to indicate a fairly targeted advertising campaign rather than focusing more on casual trips, for them to become members.

### 3.3 Most popular month

```{r echo=FALSE}
plot_3 <- ggplot(data=all_trips_cleaned, aes(fill=member_casual, x=month)) +
  geom_bar(position="dodge", stat="count") +
  scale_y_continuous(labels = scales::label_number_si()) +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(y = "trips", x = "month", fill = "type") +
  scale_fill_viridis_d()
```

```{r echo=FALSE}
plot_3
```

From the above-mentioned bar chart the following conclusions can be made :

* Trip volumes are increasing, but also there seem to be seasonal influences, with fewer trips taken in the winter months, and more in the summer months, probably due to weather conditions

* Casual trip volumes are higher than member trip volumes in the summer months, which may make sense due to the weather and vacationing

### 3.4 Most popular day of week

```{r echo=FALSE}
plot_4 <- ggplot(data=all_trips_cleaned, aes(fill=member_casual, x=factor(day_of_week))) +
  geom_bar(position="dodge", stat="count") +
  scale_y_continuous(labels = scales::label_number_si()) +
  labs(y = "trips", x = "day of week") +
  scale_fill_viridis_d()
```

```{r echo=FALSE}
plot_4
```

The above-mentioned histogram demonstrates that there are more casual than member trips on weekends, and on weekdays there are more member trips.

### 3.5 Most popular time of day

```{r echo=FALSE}
all_trips_cleaned$ToD_convert <- as.POSIXct(all_trips_cleaned$ToD,
                                            
                                            format = "%H:%M:%S")

all_trips_cleaned$by60 <- cut(all_trips_cleaned$ToD_convert, breaks ="60 mins")

circular_bar_chart_data <- all_trips_cleaned %>% 
  
  group_by(by60, member_casual) %>% 
  
  mutate(numtrips_0000s =(n()/1000)) %>% 
  
  distinct(by60, member_casual, numtrips_0000s)
```

```{r echo=FALSE}
plot_5 <- ggplot(circular_bar_chart_data) +
  
  geom_hline( aes(yintercept = y), data.frame(y = c(0:4) * 125), 
              color = "lightgrey" ) +
  
  geom_bar(
    aes( x = by60, y = numtrips_0000s, fill = member_casual ), stat="identity" ) +
      
  coord_polar(start = -0.135, direction = 1) + 
  
  ylim(-600, 500) +
  
  annotate( x = 1, y = -50, label = "00:00", geom = "text", size = 2, ) +
  
  annotate( x = 2, y = -50, label = "01:00", geom = "text", size = 2, ) +
  
  annotate( x = 3, y = -50, label = "02:00", geom = "text", size = 2, ) +
  
  annotate( x = 4, y = -50, label = "03:00", geom = "text", size = 2, ) +
  
  annotate( x = 5, y = -50, label = "04:00", geom = "text", size = 2, ) +
  
  annotate( x = 6, y = -50, label = "05:00", geom = "text", size = 2, ) +
  
  annotate( x = 7, y = -50, label = "06:00", geom = "text", size = 2, ) +

  annotate( x = 8, y = -50, label = "07:00", geom = "text", size = 2, ) +
  
  annotate( x = 9, y = -50, label = "08:00", geom = "text", size = 2, ) +
  
  annotate( x = 10, y = -50, label = "09:00", geom = "text", size = 2, ) +
  
  annotate( x = 11, y = -50, label = "10:00", geom = "text", size = 2, ) +
  
  annotate( x = 12, y = -50, label = "11:00", geom = "text", size = 2, ) +
  
  annotate( x = 13, y = -50, label = "12:00", geom = "text", size = 2, ) +
  
  annotate( x = 14, y = -50, label = "13:00", geom = "text", size = 2, ) +
  
  annotate( x = 15, y = -50, label = "14:00", geom = "text", size = 2, ) +
  
  annotate( x = 16, y = -50, label = "15:00", geom = "text", size = 2, ) +
  
  annotate( x = 17, y = -50, label = "16:00", geom = "text", size = 2, ) +
  
  annotate( x = 18, y = -50, label = "17:00", geom = "text", size = 2, ) +
  
  annotate( x = 19, y = -50, label = "18:00", geom = "text", size = 2, ) +
  
  annotate( x = 20, y = -50, label = "19:00", geom = "text", size = 2, ) +
  
  annotate( x = 21, y = -50, label = "20:00", geom = "text", size = 2, ) +
  
  annotate( x = 22, y = -50, label = "21:00", geom = "text", size = 2, ) +
  
  annotate( x = 23, y = -50, label = "22:00", geom = "text", size = 2, ) +
  
  annotate( x = 24, y = -50, label = "23:00", geom = "text", size = 2, ) +
  
  
  
  annotate( x = 24, y = 125, label = "125,000", geom = "text", size = 2.5, 
            
            angle = 15 ) +
  
  annotate( x = 24, y = 250, label = "250,000", geom = "text", size = 2.5, 
            
            angle = 15 ) +
  
  annotate( x = 24, y = 375, label = "375,000", geom = "text", size = 2.5, 
            
            angle = 15 ) +
  
  annotate( x = 24, y = 500, label = "500,000 trips", geom = "text", size = 2.5, 
            
            angle = 15 ) +

  
  scale_fill_viridis_d() +
  
  theme_light() +
  
  theme(
    axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank(), legend.position = "bottom", legend.title = element_blank(),)
```

```{r echo=FALSE}
plot_5
```

From the above-mentioned circular bar chart the following conclusions can be made :

* 5 pm is the most popular time of day for Cyclistic users, with the most common time of day being 17:19:15 for casual riders and 17:20:37 for members.

* Casual riders use the service more than members between 9 pm and 5 am. This could suggest that casual riders may be using the bike share service instead of taking taxi’s home after a night out. This is further reflected in the average time of day for casual riders which are 15:11:59, as opposed to members whose average time of day is 14:32:12.

* The significant increase of member riders using the bikes at 8 am and later on around 5 pm suggests that a lot of member riders are using the Cyclistic service for their work commutes.

* The average ride length for casual riders is 37.62 minutes, while members ride for an average of 14.39 minutes. Again this further strengthens the argument that casual riders use Cyclistic bikes for leisure purposes.

### 3.6 Most popular time of year

```{r echo=FALSE}
all_trips_cleaned$day_of_week <- ordered(all_trips_cleaned$day_of_week,
                                         levels = c("Monday", "Tuesday",
                                                    "Wednesday", "Thursday",
                                                    "Friday", "Saturday",
                                                    "Sunday"))

heatmap_data <- all_trips_cleaned %>% 
  select(YMD, day_of_week, week, year) %>% 
  group_by(YMD) %>% 
  mutate(numtrips = n()) %>% 
  distinct(YMD, .keep_all =TRUE)
```

```{r echo=FALSE}
plot_6 <- ggplot(heatmap_data, aes(x = week, y = day_of_week, fill = numtrips)) +
  scale_fill_viridis(option = "D", direction = 1, name = "Number of trips") +
  geom_tile(colour = "white", na.rm = FALSE) +
  facet_wrap("year", ncol =1) +
  scale_y_discrete(limits = rev) +
  scale_x_discrete(expand = c(0,0), breaks = seq(1, 52, length = 12),
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  theme_light()+
  theme(axis.title = element_blank())
```

```{r echo=FALSE}
plot_6
```

The above-mentioned heat map demonstrates that the summer months are the most popular time of year for Cyclistic and the weekends are the most popular time of the week. Indicating that the majority of users are riding for leisure purposes. In addition, the increase in trips during the summer of 2021 in comparison to the summer months of 2020 depicts the change in public behaviour due to the COVID-19 pandemic.

```{r echo=FALSE}
heatmap_data_member_casual <- all_trips_cleaned %>% 
  select(YMD, day_of_week, week, year, member_casual) %>% 
  group_by(member_casual, YMD) %>% 
  mutate(numtrips = n()) %>% 
  distinct(member_casual, YMD, .keep_all =TRUE)

member_heatmap <- heatmap_data_member_casual %>% 
  filter(member_casual == "member")

casual_heatmap <- heatmap_data_member_casual %>% 
  filter(member_casual == "casual")
```

```{r echo=FALSE}
plot_6_member <- ggplot(member_heatmap, aes(x = week, y = day_of_week, fill = numtrips)) +
  scale_fill_viridis(option = "D", direction = 1, name = "Number of trips") +
  geom_tile(colour = "white", na.rm = FALSE) +
  facet_wrap("year", ncol =1) +
  scale_y_discrete(limits = rev) +
  scale_x_discrete(expand = c(0,0), breaks = seq(1, 52, length = 12),
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  theme_light()+
  theme(axis.title = element_blank()) +
  labs(title = "Member Riders")
```

```{r echo=FALSE}
plot_6_casual <- ggplot(casual_heatmap, aes(x = week, y = day_of_week, fill = numtrips)) +
  scale_fill_viridis(option = "D", direction = 1, name = "Number of trips") +
  geom_tile(colour = "white", na.rm = FALSE) +
  facet_wrap("year", ncol =1) +
  scale_y_discrete(limits = rev) +
  scale_x_discrete(expand = c(0,0), breaks = seq(1, 52, length = 12),
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  theme_light()+
  theme(axis.title = element_blank()) +
  labs(title = "Casual Riders")
```

```{r echo=FALSE}
plot_6_member_casual <- ggarrange(plot_6_member, plot_6_casual, ncol = 1, nrow = 2,
                     common.legend = TRUE, legend = "right")
```

```{r echo=FALSE}
plot_6_member_casual
```

From the above-mentioned heat map the following conclusions can be made :

* Members use the Cyclistic service more consistently throughout the year in comparison to casual riders who tend to only use the bike service during the summer months on the weekend.

* The fact that the most common weekday for members was Wednesday while that for the casual riders was Saturday, which similar to the other Sections also indicates members use their Cyclistic service for their work commute and other daily activities, unlike casual riders who are using the Cyclistic service for predominately leisure purposes.

# 4. Recommendations

As the primary objective of these recommendations concluded from the insights of this case study is to design a marketing campaign that should not be focused on encouraging new customers to use the Cyclistic bike service but instead focus on encouraging casual riders to convert to annual memberships.

The two possible recommendations for the marketing campaign of Cyclistic bike-sharing company are as follows:

***1. Digital campaign demonstrating how Cyclistic’s bike service fits in a Chicago local’s everyday life***

The above analysis demonstrates that the peak time of day for both casual and member riders was 5 pm. Even though casual riders use the bikes slightly more than members during the evening and early hours of the morning, the most popular time of day for both is still during the day with a slight peak during work commute times. 

This shows that there are casual riders using the Cyclistic service for their work commute without investing in an annual membership. To encourage casual riders to convert to members, a digital campaign that encourages Chicago locals to see how Cyclistic’s services fit into their everyday life would be beneficial.

In addition to this, it seems that the marketing campaign should try to seek out the customers that are using the bicycles outside of the summer months, where it is more likely that these are really taken for incidental rather than regular use and the customers who are taking the bicycles on weekdays, as they are more likely to use them on a regular basis than the ones taken on weekends.

***2. Regular and periodical notifications/ email reminders for casual riders to observe the various benefit of converting to annual memberships***

Similar to the insight noted above, for Chicago, locals to convert to an annual membership from single-ride passes or full-day, phone notifications and email reminders should be used to remind regular casual riders (most likely Chicago locals) of various benefits that they can redeem from purchasing an annual membership rather continuing their regular casual trips and paying on a daily/ride basis.

In addition to this, it seems that the marketing campaign should try to seek out the customers that are using the bicycles outside of the summer months, where it is more likely that these are really taken for incidental rather than regular use and the customers who are taking the bicycles on weekdays, as they are more likely to use them on a regular basis than the ones taken on weekends.



